---
title: "breast_bladder_thyroid_preprocessing"
author: "Thais Minet"
date: "3/3/2018"
output: html_document
---

# Preprocessing of RNA-seq data

***
> p = paran(data.log@data, iterations = 500)

Using eigendecomposition of correlation matrix.
Computing: 10%  20%  30%  40%  50%  60%  70%  80%  90%  100%


Results of Horn's Parallel Analysis for component retention
500 iterations, using the mean estimate

-------------------------------------------------- 
Component   Adjusted    Unadjusted    Estimated 
            Eigenvalue  Eigenvalue    Bias 
-------------------------------------------------- 
1         776.43023  776.879466      0.449227
2          16.617828   17.060942      0.443114
3           8.873398    9.312026      0.438628
4           6.466199    6.901024      0.434824
5           3.831106    4.262450      0.431343
6           1.155209    1.583063      0.427853
-------------------------------------------------- 

Adjusted eigenvalues > 1 indicate dimensions to retain.
(6 components retained)
***


### Load Utility functions
```{r}
source("utility.R")
library("paran")
```


### Data preprocessing

Load all of the unnormalized data into `data`.
```{r}
file = "data/all-unnorm.csv"
data = read.csv(file, header = TRUE, row.names = 1)
rm(file)
print(dim(data))
```

```{r}
data[1:5, 1:2]
```

Load the data into the S4 `rnaObj`.
```{r}
data = rnaObj(data = data)
```

Now we initialize the object choosing the tissues we want to analyze and prepare for the neural network. The tissues available are `bladder`, `breast`, `cervix`, `colon`, `liver`, `prostate`, `stomach`, `thyroid` and `uterus`. To select only a subset of the tissues provide a vector to the initialization function that specifies which tissues to keep e.g. `initialize(data, keep.tissue = c("bladder", "breast", "thyroid"))`. By default no transformation are applied to the data when it is initialized. To apply a log transformation set `log = TRUE`. To apply z-score set `normalize = TRUE`. To quantile normalize set `qn = TRUE`.
```{r}
tissues = c("breast", "prostate", "thyroid")

data = initialize(data, keep.tissue = tissues)
data.log = initialize(data, keep.tissue = tissues, log = TRUE)
data.norm = initialize(data, keep.tissue = tissues, log = TRUE, normalize = TRUE)
data.qn = initialize(data, keep.tissue = tissues, log = TRUE, qn = TRUE)

# data = initialize(data)
# data.log = initialize(data, log = TRUE)
# data.norm = initialize(data, log = TRUE, normalize = TRUE)
# data.qn = initialize(data, log = TRUE, qn = TRUE)
```

Display how many of each tissue we have.
```{r}
print(data@counts)
```


### PCA & Scatterplot
Compute PCA of each object.
```{r}
data = doPCA(data)
data.log = doPCA(data.log)
data.norm = doPCA(data.norm)
data.qn = doPCA(data.qn)
```


Do a scatter plot of the data to visualize.
```{r}
scatterPlot(data, title = "Unnormalized data")
scatterPlot(data.log, title = "Unnormalized log transformed data")
scatterPlot(data.norm, title = "Unnormalized log transformed and z-scored data")
scatterPlot(data.qn, title = "Unnormalized log transformed and quantile normalized data")
```

Do combat things
```{r}
annotation = data.log@meta
rownames(annoation) = colnames(data.log@data)

batch = data.log@meta$study
design = model.matrix(~1, data = data.log@meta)
combat = ComBat(dat = data.log@data, batch = batch, mod = design, par.prior = TRUE)

pheatmap(cor(data.log@data, method='spearman'), main="after combat", show_rownames = FALSE, show_colnames = FALSE, annotation = annotation, annotation_row = annotation, annotation_legend = TRUE)
pheatmap(cor(combat, method='spearman'), main="after combat", show_rownames = FALSE, show_colnames = FALSE, annotation = annotation, annotation_row = annotation, annotation_legend = TRUE)

combat.pc = fast.prcomp(t(combat), center = FALSE, scale = FALSE)

combat.pc.scores = data.frame(combat.pc$x[ ,1:100])
combat.pc.load = data.frame(combat.pc$rotation[ , 1:100])
combat.pc.percentage = (100*(combat.pc$sdev)^2 / sum(combat.pc$sdev^2))[1:100]

combat.pc.meta = combat.pc.scores
combat.pc.meta[, 'study'] = data.log@study
combat.pc.meta[, 'tissue'] = data.log@tissue
combat.pc.meta = arrange.vars(combat.pc.meta, c("study"=1, "tissue"=2))

tissue = factor(data.log@tissue)
study = factor(data.log@study)

ggplot(combat.pc.scores, aes(x = PC1, y = PC2, shape = study, colour = tissue)) + 
              geom_point(size = 2) + scale_color_hue(l = 55) + 
              theme_bw() + 
              guides(colour = guide_legend(order = 2), shape = guide_legend(order = 2)) +
              ggtitle("Unnormalized log transformed with ComBat") + 
              theme(plot.title = element_text(hjust = 0.5, lineheight=.8, face="bold"))
```

### Split data into GTEX and TCGA batches for input to neural network.
```{r}
log.gtex = data.log@pca.meta[grep("GTEX", rownames(data.log@pca.meta)), ]
log.tcga = data.log@pca.meta[grep("TCGA", rownames(data.log@pca.meta)), ]

norm.gtex = data.norm@pca.meta[grep("GTEX", rownames(data.norm@pca.meta)), ]
norm.tcga = data.norm@pca.meta[grep("TCGA", rownames(data.norm@pca.meta)), ]

qn.gtex = data.qn@pca.meta[grep("GTEX", rownames(data.qn@pca.meta)), ]
qn.tcga = data.qn@pca.meta[grep("TCGA", rownames(data.qn@pca.meta)), ]
```

### Write PCA data to file for input to neural network
Write respective batches
```{r}
write.csv(log.gtex[, 1:22], file = "data/processed/unnorm-log-20PC-GTEX-breast-prostate-thyroid.csv", row.names = TRUE)
write.csv(log.tcga[, 1:22], file = "data/processed/unnorm-log-20PC-TCGA-breast-prostate-thyroid.csv", row.names = TRUE)

write.csv(norm.gtex[, 1:22], file = "data/processed/unnorm-norm-20PC-GTEX-breast-prostate-thyroid.csv", row.names = TRUE)
write.csv(norm.tcga[, 1:22], file = "data/processed/unnorm-norm-20PC-TCGA-breast-prostate-thyroid.csv", row.names = TRUE)

write.csv(qn.gtex[, 1:22], file = "data/processed/unnorm-qn-20PC-GTEX-breast-prostate-thyroid.csv", row.names = TRUE)
write.csv(qn.tcga[, 1:22], file = "data/processed/unnorm-qn-20PC-TCGA-breast-prostate-thyroid.csv", row.names = TRUE)

rm(log.gtex, log.tcga, norm.gtex, norm.tcga, qn.gtex, qn.tcga)
```


Write complete data frame
```{r}
write.csv(data@pca.scores[, 1:22], file = "data/processed/unnorm-20PC-df-breast-prostate-thyroid.csv", row.names = TRUE)
write.csv(data.log@pca.scores[, 1:22], file = "data/processed/unnorm-log-20PC-df-breast-prostate-thyroid.csv", row.names = TRUE)
write.csv(data.norm@pca.scores[, 1:22], file = "data/processed/unnorm-norm-20PC-df-breast-prostate-thyroid.csv", row.names = TRUE)
write.csv(data.qn@pca.scores[, 1:22], file = "data/processed/unnorm-qn-20PC-df-breast-prostate-thyroid.csv", row.names = TRUE)
```